### 2019-12-06 paper3/article.tex
Modified reference style.

Recent advances in colloidal science have generated a growing interest in the fabrication of functional materials, especially those possessing photonic band gaps \cite{PRL1990, PBM1999, ZnO2003, colloidal2003, Shen_2016AS}. Colloids --  particles or droplets of (sub-)micron scale -- are manipulated with high precision to self-organize into controlled patterns, which then form a library of basic building blocks for more complex structures \cite{Sacanna_Pine_2011}. Conceivably, such direct assembly is also the most efficient mechanism for material synthesis. Colloidal self-assembly has thus become a promising scheme to engineer novel materials, mimicking the machinery of nature \cite{vanBlaaderen}.

Currently, there are many strategies to synthesize colloidal building blocks, e.g.\ creating a short-range depletion potential in a microwell \cite{Meng2010}, exploiting shape/surface anisotropy of the colloids \cite{Sacanna_Pine_2011, Evers2016}, using patterned substrates as templates \cite{Yin2001}, or even jamming droplets with a micro-mixer \cite{Joshua_2017PRL}. Among these, \textit{flow-assisted} self-assembly seems especially appealing, since the microfluidic droplets are driven by an external flow rather than Brownian motions, reducing the range of assembly time from days down to seconds (c.f.\ the experiment of McMullen \etal \cite{McMullen2018} as an example). With the potential of being optimized and applied in parallel, microfluidics appears to be the fastest pathway towards photonic material generation.

Following this approach,   \cite{Shen_2016AS} recently demonstrated the generation and self-assembly of droplet strings into a rich variety of (non-)compact structures, including chains, triangles, diamonds, crosses, \etc .\ in a simple microfluidic device (see Fig.\ \ref{fig: chip}). Unlike most previous studies \cite{Cui2004,Beatus2006,Janssen2012,Uspal2013,Desreumaux,zhu_gallaire_2016}, where droplets under strong confinement interact via dipolar flows \cite{q2d_Beatus,Diamant}, these droplets are smaller than the smallest dimension of the microfluidic channel. Specifically, the droplets are tens to hundreds microns in diameter, while the channel height is about three times larger. As a consequence, typical quasi two-dimensional ($q2D$) arguments do not apply, and the dynamics is fully three-dimensional ($3D$). In the latter case, the complete Navier-Stokes (or Stokes) equations have to be solved to obtain a correct physical understanding of the system,\footnote{A notable exception is the dynamics of two droplets in unbounded linear flows solved analytically by Batchelor and Green \cite{batchelor_green_1972}.} wherein prior simulations mainly focus on the deformation and migration of a single drop, or an evenly spaced array of drops \cite{coulliette_pozrikidis_1998,Griggs2007,Janssen_Anderson_2007}. Interestingly, assuming a tunable far-field dipolar interaction,   \cite{Shen_2016AS} could however model the droplet motions up to a ``semiquantitative'' level, reproducing the sophisticated self-assembly observed in experiments. This apparent contradiction motivates us to pursue a detailed numerical study of the relatively fast droplet self-assembly in order to investigate the interactions in their microfluidic channel.

@@

The dynamics of two immiscible, Newtonian fluids in incompressible flows is governed by the Navier-Stokes equations
\begin{subequations}
 \begin{equation}
   \nabla \cdot {\bm u} = 0,
  \label{div free}
 \end{equation}
 \begin{equation}
      \frac{\partial {\bm u}}{\partial t} + {\bm u} \cdot \nabla {\bm u} = \frac{1}{\rho_iRe} \bigg(-\nabla p + \nabla \cdot \big[ \mu_i ( \nabla {\bm u} + \nabla {\bm u}^T ) \big] \bigg) + \frac{{\bm g}}{Fr},
  \label{NS}
 \end{equation}
\end{subequations}
where ${\bm u}$, $p$, $\bm{g}$, $\rho_i$ and $\mu_i$ denote the non-dimensional velocity, pressure, unit gravitational vector, density, and dynamic viscosity, respectively. Eq.\ \eqref{NS} is written for each fluid component, $i$ ($=1$ for the carrier fluid, 2 for the droplet), requiring a pressure boundary condition across the fluid interface
\begin{equation}
  p_2 - p_1 = \frac{\kappa}{Ca},
  \label{p laplace}
\end{equation}
with $\kappa$ being the mean curvature (the contribution due to the viscosity difference is neglected assuming matching viscosity \cite{Batchelor}).

@@

Droplets suspended in an ambient fluid dissolved by surfactant molecules typically experience an attractive depletion force. The first model to describe such interaction was proposed by Asakura and Oosawa \cite{Asakura_Oosawa}, who assumed the surfactant micelles to be non-interacting hard spheres. As sketched in Fig.\ \ref{fig: depl theory}, a suspension of such small spheres around the large colloidal droplets creates an osmotic pressure on the droplet surface. When the distance between two droplets is less than the diameter of the surfactant micelles, the pressure defect due to the exclusion of the micelles pulls the drops even closer, corresponding to an attractive force. Integrating this force with respect to the inter-droplet distance $r$ leads to a potential energy
\begin{equation}
    U(r)=
    \begin{cases}
        \infty & \textrm{if} \quad r \leqslant 2R \\
        - p_{os}V_{ex} & \textrm{if} \quad 2R < r \leqslant 2R+2r_s  \\
        0 \quad & \textrm{otherwise}, \\
    \end{cases}
    \label{pot}
\end{equation}
where $V_{ex}$ is the excluded volume and $p_{os}$ is the osmotic pressure. For spherical droplets, $V_{ex}$ can be calculated analytically
\begin{equation}
    V_{ex}(r) = \frac{4\pi (R+r_s)^3}{3}\bigg[ 1- \frac{3r}{4(R+r_s)}+\frac{r^3}{16(R+r_s)^3} \bigg],
    \label{pot V}
\end{equation}
where $R$ and $r_s$ are, respectively, the radii of the big and small spheres. The osmotic pressure is given as
\begin{equation}
    p_{os} = nkT,
    \label{pot P}
\end{equation}
where $n$ is the number density of the small spheres, $k$ the Boltzmann constant, and $T$ the temperature. The negative sign in Eq.\ \eqref{pot} corresponds to the tendency of the system to reduce its potential energy as the overlap increases. This is equivalent to increasing the total entropy of the small spheres \cite{Melby_PRL_2007}, and it provides a physical description of the depletion force even when the droplets are deformable, or when $p_{os}$ cannot be expressed by the van't Hoff's formula (Eq.\ \eqref{pot P}) \cite{Asakura_Oosawa}.

Based on the above theory, we consider the depletion potential in the hydrodynamic model by adding an osmotic pressure, $p'(r')$, near the overlap region ($r'$ being the normalized distance to the droplet surface, see Fig.\ \ref{fig: depl theory}). Specifically, we write $p'$ as a Taylor-series expansion from $r'=r_s$
\begin{equation}
    p'(r'/r_s) = p'(1) + \bigg(\frac{r'}{r_s}-1 \bigg) \frac{\partial{p'}}{\partial{r'/r_s}},
    \label{pot P linear}
\end{equation}
with a constant $\partial{p'}/\partial{(r'/r_s)}$ satisfying the original depletion force acting on one droplet, i.e.
\begin{equation}
    -p_{os}A_{ex} = \int_\Omega \big(p'(1) - p'(r'/r_s)\big)dS,
    \label{pot P balance}
\end{equation}
where $A_{ex}$ is the effective area of the overlap $\Omega$. Note that, in this formulation, we do not require $p_{os}$ to be a thermodynamic pressure; instead, its strength can be defined by a non-dimensional number
\begin{equation}
    \Pi = \frac{p_{os}}{\kappa /Ca},
    \label{Pi}
\end{equation}
which normalizes $p_{os}$ by the Laplace pressure due to surface tension. Doing so, the osmotic pressure varies continuously within the overlap and depends linearly on $r'$. An expansion of the osmotic pressure with the distance corresponds to a gradient of the micelle concentration near the gap; and if the micelle is much smaller than the droplet, as it is in the microfluidic device of interest \cite{Shen_2016AS}, the gradient will be very sharp. Conversely, when the distance to the surface varies slowly, such as in the gap of a \xx{squeezed} droplet and a flat wall, a uniform pressure will be recovered. In essence, what we propose here is a method to model the osmotic pressure as an equivalent mechanical pressure such that a favorable pressure gradient from the center of the overlap region generates an outflow, pulling the droplets towards each other. This is clearly illustrated in Fig.\ \ref{fig: depl flow}, where two droplets approach each other due to the locally induced depletion flow.

@@

The above governing equations are solved numerically using the interface-correction level set/ghost fluid method (ICLS/GFM) \cite{ICLS}. The equations are discretized in space by the finite volume method, and integrated in time using the second-order Adam-Bashforth scheme. Combining several computational techniques, including the fast pressure correction method \cite{Dodd_JCP_2014}, the GFM \cite{Fedkiw_JCP_1999}, and the fast Fourier transform (FFT) \cite{Schumann_JCP_1988}, discontinuous quantities are treated sharply at high efficiency. For the detailed algorithm and validations, we refer to Ge \etal \cite{ICLS}. The source code is also publicly available on GitHub \cite{icls_code}.

@@

The simplest case of droplet self-assembly is identical drops approaching in quiescent flows, i.e.\ no external flow motions. In such a case, the remaining fluid parameters can be grouped into two non-dimensional numbers, i) the Laplace number $La = \tilde{\sigma} \tilde{\rho}_1 (2\tilde{R})/\tilde{\mu}_1^2$, where $\tilde{R}$ is the droplet radius; and ii) the osmotic-to-Laplace pressure ratio $\Pi=p_{os}/p$. \footnote[3]{The density and viscosity ratios between the drop and the carrier fluids are assumed to be unity. See Appendix \ref{app:numerics} for the detailed numerical setup.} $La$ relates surface tension to the viscous stress, which can also be expressed as $La = Re/Ca$. $\Pi$ indicates the magnitude of the uniform osmotic pressure $p_{os}$ due to depletion of the surfactant micelles, scaled by a reference Laplace pressure $p$ due to surface tension (cf.\ Eq.\ \eqref{Pi}). In the following, we assume $La = 2000$ and $\Pi = 1/8$ or $1/2$, corresponding to the limit $Ca \ll Re \ll 1 $ and conditions above the critical micelle concentrations (CMC, see dimensional analysis below) as in the experiments \cite{Shen_2016AS}. 
%This is reasonable since Re $\sim 10^{-3}$ and Ca $\sim 10^{-6}$ (see \cite{shen_thesis}, pp 63-64), and the osmotic pressure at surfactant concentrations above CMC is of similar order-of-magnitude to the Laplace pressure (see dimensional values below). The same setup is used throughout the section, except that the domain size is adjusted to accommodate more droplets when necessary.

@@

We note that $T_\pi$ is not a physical time scale (it is dimensionless). One possible definition for the depletion time scale is $\tilde{\tau}_\pi = \tilde{r}_s \tilde{\mu}_1 /(\tilde{R} \tilde{p}_{os})$, which can be rewritten as $T_\pi Ca \tilde{\tau}$, with a convection time scale $\tilde{\tau} =2\tilde{R}/\tilde{U}$ in the range of $0.1$ to $1$ s typically. By substitution of usual values of colloidal systems, e.g. $\tilde{r}_s = 1$ nm, $\tilde{\mu}_1 = 10^{-3}$ kg/m-s, $\tilde{R} = 10$ $\mu$m, and $\tilde{p}_{os} = 100$ Pa (corresponding to surfactant micelles concentration of 5 CMC, see Shen \cite{shen_thesis}, p112), the estimated time scale is 1 ns. Although $\tilde{\tau}_\pi$ can be amplified by increasing the viscosity of the suspending fluid or reducing the micelles concentration, its magnitude is so small that the approaching can be considered instantaneous. Therefore, in practical microfluidic devices such as those in   \cite{Shen_2016AS}, one cannot expect to detect the dynamical approaching process due to depletion forces. The droplets will appear either bound or separated, depending on the surfactant concentration and flow conditions.

@@

To further demonstrate the effect of the depletion force, we ``virtually'' assemble four to ten droplets under various initial configurations to form stable clusters as illustrated in Fig.\ \ref{fig: sph pack}. These clusters can be either $2D$ or $3D$, exhibiting different levels/kinds of symmetry. In our simulations, the shape of the cluster is solely determined by the initial droplet arrangement, in the absence of any disturbance or other driving forces. Permitting disturbances, such as vibrations or thermal noises, would eventually lead to the formation of ``rigid clusters", i.e.\ clusters that cannot be reshaped by a small amount of inter-droplet displacement. Analytically, the number of possible rigid clusters grows rapidly with the number of droplets ($N$). For example, there is only one possible rigid cluster for $N = 4$, while there are 259 possibilities for $N = 10$, for packing of $3N-6$ contacts  \cite{Holmes-Cerfon}. For brevity, we only illustrate two examples for $N = 4$ (including one planar cluster) and one example for $N = 5$ to 10 in Fig. \ref{fig: sph pack}.

We remind that the specific coordinates of the sphere packings bear no more significance than other possibilities in our simulations. They are arbitrarily chosen to illustrate the self-assembly due to the near field attraction. This is not the case for trapped equilibrium clusters, where less symmetric geometries are found to be favored by the entropic depletion force \cite{Meng2010,Klein2018}. \xx{Nor is it similar to colloidal particles interacting via short-range attracitve, long-range repulsive potentials, where complex phase transitions emerge depending on the competition of the interactions \cite{salr14,salr18}.} Here, the self-assembly is microfluidic-based, driven by the hydrodynamics rather than the minimization of free energy over long periods. We examine the effect of the flow next.

@@

When the droplets are carried by an external flow, their interactions are undoubtedly affected by the flow conditions, droplet-flow interactions and flow-induced droplet-droplet interactions \cite{Fouxon_2017}. To study these additional effects, we consider an elementary flow field, the wall-bounded simple shear flow, defined as $(u,v,w)=(0, \dot{\gamma}z,0)$ for $z \in [-L_z/2,L_z/2]$. Here, $v$ is the only non-zero velocity component, its magnitude varies linearly with the $z$ coordinate, and $\dot{\gamma}$ is the shear rate (see Fig.\  \ref{fig: shell}a). The presence of droplets will locally modify this flow field, which we sustain by enforcing opposite motions of two moving plates at $z= \pm L_z/2$.

Dating back to Taylor \cite{Taylor501}, the deformation and motion of single or multiple droplet(s)/particle(s) have been studied extensively in simple shear flows \cite{lin_lee_sather_1970,batchelor_green_1972,batchelor_green_1972b,Zinchenko1983,Zinchenko1984,zurita-gotor_2007}. In the case of spherical particles/droplets, previously identified interaction modes include closed trajectories (particles rotate around each other in the vorticity plane), open-and-symmetric trajectories (particles return to their original $z$ positions after passing each other), and swapping trajectories (particles exchange the $z$ position after a binary encounter). The first two modes are generic features of a dilute suspension of particles or non-deforming/non-coalescing droplets, while the last mode arises when the particles are under relatively large geometric confinement. Conceivably, adding a near-field depletion force shall not alter these three modes as it is only activated at nearly touching, when the particles are already in a bound pair. What is yet to be explored, however, is when the gap between the two confining plates is smaller than the sum of the particle diameters (i.e.\ $L_z/D<2$), thus disabling the occurrence of the Batchelor-Green type of closed orbit.

@@

Fig.\ \ref{fig: shell}(a) illustrates various initial conditions of two touching droplets, corresponding to the nine cases listed in Table \ref{tbl: cases}, at $L_z/D=1.5$. Specifically, the initial positions of the second droplet is given in the spherical coordinates centered at the first one, where all cases are located in one quadrant-sphere as we do not distinguish between the two droplets (the rest are equivalent due to symmetry). Contrary to the less confined conditions where a bound pair would rotate indefinitely relative to each other under shear \cite{batchelor_green_1972,Zinchenko1984}, Fig.\ \ref{fig: shell} shows that the droplets tend to reside in the vorticity plane (i.e.\ the $yz$ plane) with a stable polar angle $\theta_\infty \approx 79$ deg. This is true even if the two droplets are not initially in the same vorticity plane (case 2,3,5,6,9), or if the second droplet is in the lower hemisphere relative to the first one (case 7,8,9). Particularly, in cases 8 and 9, the droplet pair first rotates clockwise, then slides along the wall, before finally reaching the stable orientation (see ESI Video 1). The anomalous trajectory is a clear evidence of the influence of the walls, which, together with the attractive depletion force, break the symmetry of the droplet binary interactions.

@@

\xx{In the following, clusters of three or four droplets are initialized to be in contact ($N-1$ contacts for $N$ droplets) and are released into different regions of a microfluidic channel to study their transport behavior. The production of these droplets is omitted, as the step-emulsifier is typically much smaller than the size of the channel, allowing for separation of the two processes \cite{shen_thesis,step-emuls}. We note that, although the droplets are already in a cluster initially, their relative rearrangement is still important as it determines the cluster morphology in the final state. The latter results primarily from the droplet-flow interaction and has direct consequence on the photonic properties of the droplet lattice, as we will discuss in details below.}

@@

As a control case, we simulate three droplets initially located near the bottom of the channel without any depletion forces. This is shown in the first column of Fig.\ \ref{fig: poi results}a, where the top and bottom panels illustrate the top and side views of the channel (see Fig.\ \ref{fig: chip}). The two snapshots are separated by $13.5$ convection time units (i.e.\ $\tilde{\tau}$). Clearly, the droplets quickly scatter as carried by the flow. In a previous work, we theoretically predicted the emergence of singlets and pairs of a dilute particle suspension due to weak particle-particle interactions \cite{Fouxon_2017}. This example illustrates the separation of a droplet cluster, enhanced by their initial proximity already at $N=3$, supporting our theoretical predictions.

@@

The above results confirm that the chain-like structure is indeed the predominant configuration of droplets bound by short-range depletion forces in the Poiseuille flow. Experimentally, this corresponds to strong diluting flows at the channel inlet, where long droplet strings are also observed further downstream (see Shen \cite{shen_thesis}, p137). More importantly, our simulations suggest that aligning of the droplets is a $3D$ shear-induced effect mediated by the confinement. The cross-stream migration of the droplets happens within a much shorter time span than any tangential rearrangement due to the dipolar interactions (cf.\ Diamant \cite{Diamant} and Fouxon \etal \cite{Fouxon_2017}). This is one key difference between our $3D$ microfluidic channel and other $q2D$ devices.

>>>>

Recent advances in colloidal science have generated a growing interest in the fabrication of functional materials, especially those possessing photonic band gaps \citep{PRL1990, PBM1999, ZnO2003, colloidal2003, Shen_2016AS}. Colloids --  particles or droplets of (sub-)micron scale -- are manipulated with high precision to self-organize into controlled patterns, which then form a library of basic building blocks for more complex structures \citep{Sacanna_Pine_2011}. Conceivably, such direct assembly is also the most efficient mechanism for material synthesis. Colloidal self-assembly has thus become a promising scheme to engineer novel materials, mimicking the machinery of nature \citep{vanBlaaderen}.

Currently, there are many strategies to synthesize colloidal building blocks, e.g.\ creating a short-range depletion potential in a microwell \citep{Meng2010}, exploiting shape/surface anisotropy of the colloids \citep{Sacanna_Pine_2011, Evers2016}, using patterned substrates as templates \citep{Yin2001}, or even jamming droplets with a micro-mixer \citep{Joshua_2017PRL}. Among these, \textit{flow-assisted} self-assembly seems especially appealing, since the microfluidic droplets are driven by an external flow rather than Brownian motions, reducing the range of assembly time from days down to seconds (c.f.\ the experiment of \citep{McMullen2018} as an example). With the potential of being optimized and applied in parallel, microfluidics appears to be the fastest pathway towards photonic material generation.

Following this approach,   \cite{Shen_2016AS} recently demonstrated the generation and self-assembly of droplet strings into a rich variety of (non-)compact structures, including chains, triangles, diamonds, crosses, \etc .\ in a simple microfluidic device (see Fig.\ \ref{fig: chip}). Unlike most previous studies \citep{Cui2004,Beatus2006,Janssen2012,Uspal2013,Desreumaux,zhu_gallaire_2016}, where droplets under strong confinement interact via dipolar flows \citep{q2d_Beatus,Diamant}, these droplets are smaller than the smallest dimension of the microfluidic channel. Specifically, the droplets are tens to hundreds microns in diameter, while the channel height is about three times larger. As a consequence, typical quasi two-dimensional ($q2D$) arguments do not apply, and the dynamics is fully three-dimensional ($3D$). In the latter case, the complete Navier-Stokes (or Stokes) equations have to be solved to obtain a correct physical understanding of the system,\footnote{A notable exception is the dynamics of two droplets in unbounded linear flows solved analytically by \cite{batchelor_green_1972}.} wherein prior simulations mainly focus on the deformation and migration of a single drop, or an evenly spaced array of drops \citep{coulliette_pozrikidis_1998,Griggs2007,Janssen_Anderson_2007}. Interestingly, assuming a tunable far-field dipolar interaction,   \cite{Shen_2016AS} could however model the droplet motions up to a ``semiquantitative'' level, reproducing the sophisticated self-assembly observed in experiments. This apparent contradiction motivates us to pursue a detailed numerical study of the relatively fast droplet self-assembly in order to investigate the interactions in their microfluidic channel.

@@

The dynamics of two immiscible, Newtonian fluids in incompressible flows is governed by the Navier-Stokes equations
\begin{subequations}
 \begin{equation}
   \nabla \cdot {\bm u} = 0,
  \label{div free}
 \end{equation}
 \begin{equation}
      \frac{\partial {\bm u}}{\partial t} + {\bm u} \cdot \nabla {\bm u} = \frac{1}{\rho_iRe} \bigg(-\nabla p + \nabla \cdot \big[ \mu_i ( \nabla {\bm u} + \nabla {\bm u}^T ) \big] \bigg) + \frac{{\bm g}}{Fr},
  \label{NS}
 \end{equation}
\end{subequations}
where ${\bm u}$, $p$, $\bm{g}$, $\rho_i$ and $\mu_i$ denote the non-dimensional velocity, pressure, unit gravitational vector, density, and dynamic viscosity, respectively. Eq.\ \eqref{NS} is written for each fluid component, $i$ ($=1$ for the carrier fluid, 2 for the droplet), requiring a pressure boundary condition across the fluid interface
\begin{equation}
  p_2 - p_1 = \frac{\kappa}{Ca},
  \label{p laplace}
\end{equation}
with $\kappa$ being the mean curvature (the contribution due to the viscosity difference is neglected assuming matching viscosity \citep{Batchelor}).

@@

Droplets suspended in an ambient fluid dissolved by surfactant molecules typically experience an attractive depletion force. The first model to describe such interaction was proposed by  \cite{Asakura_Oosawa}, who assumed the surfactant micelles to be non-interacting hard spheres. As sketched in Fig.\ \ref{fig: depl theory}, a suspension of such small spheres around the large colloidal droplets creates an osmotic pressure on the droplet surface. When the distance between two droplets is less than the diameter of the surfactant micelles, the pressure defect due to the exclusion of the micelles pulls the drops even closer, corresponding to an attractive force. Integrating this force with respect to the inter-droplet distance $r$ leads to a potential energy
\begin{equation}
    U(r)=
    \begin{cases}
        \infty & \textrm{if} \quad r \leqslant 2R \\
        - p_{os}V_{ex} & \textrm{if} \quad 2R < r \leqslant 2R+2r_s  \\
        0 \quad & \textrm{otherwise}, \\
    \end{cases}
    \label{pot}
\end{equation}
where $V_{ex}$ is the excluded volume and $p_{os}$ is the osmotic pressure. For spherical droplets, $V_{ex}$ can be calculated analytically
\begin{equation}
    V_{ex}(r) = \frac{4\pi (R+r_s)^3}{3}\bigg[ 1- \frac{3r}{4(R+r_s)}+\frac{r^3}{16(R+r_s)^3} \bigg],
    \label{pot V}
\end{equation}
where $R$ and $r_s$ are, respectively, the radii of the big and small spheres. The osmotic pressure is given as
\begin{equation}
    p_{os} = nkT,
    \label{pot P}
\end{equation}
where $n$ is the number density of the small spheres, $k$ the Boltzmann constant, and $T$ the temperature. The negative sign in Eq.\ \eqref{pot} corresponds to the tendency of the system to reduce its potential energy as the overlap increases. This is equivalent to increasing the total entropy of the small spheres \citep{Melby_PRL_2007}, and it provides a physical description of the depletion force even when the droplets are deformable, or when $p_{os}$ cannot be expressed by the van't Hoff's formula (Eq.\ \eqref{pot P}) \citep{Asakura_Oosawa}.

Based on the above theory, we consider the depletion potential in the hydrodynamic model by adding an osmotic pressure, $p'(r')$, near the overlap region ($r'$ being the normalized distance to the droplet surface, see Fig.\ \ref{fig: depl theory}). Specifically, we write $p'$ as a Taylor-series expansion from $r'=r_s$
\begin{equation}
    p'(r'/r_s) = p'(1) + \bigg(\frac{r'}{r_s}-1 \bigg) \frac{\partial{p'}}{\partial{r'/r_s}},
    \label{pot P linear}
\end{equation}
with a constant $\partial{p'}/\partial{(r'/r_s)}$ satisfying the original depletion force acting on one droplet, i.e.
\begin{equation}
    -p_{os}A_{ex} = \int_\Omega \big(p'(1) - p'(r'/r_s)\big)dS,
    \label{pot P balance}
\end{equation}
where $A_{ex}$ is the effective area of the overlap $\Omega$. Note that, in this formulation, we do not require $p_{os}$ to be a thermodynamic pressure; instead, its strength can be defined by a non-dimensional number
\begin{equation}
    \Pi = \frac{p_{os}}{\kappa /Ca},
    \label{Pi}
\end{equation}
which normalizes $p_{os}$ by the Laplace pressure due to surface tension. Doing so, the osmotic pressure varies continuously within the overlap and depends linearly on $r'$. An expansion of the osmotic pressure with the distance corresponds to a gradient of the micelle concentration near the gap; and if the micelle is much smaller than the droplet, as it is in the microfluidic device of interest \citep{Shen_2016AS}, the gradient will be very sharp. Conversely, when the distance to the surface varies slowly, such as in the gap of a \xx{squeezed} droplet and a flat wall, a uniform pressure will be recovered. In essence, what we propose here is a method to model the osmotic pressure as an equivalent mechanical pressure such that a favorable pressure gradient from the center of the overlap region generates an outflow, pulling the droplets towards each other. This is clearly illustrated in Fig.\ \ref{fig: depl flow}, where two droplets approach each other due to the locally induced depletion flow.

@@

The above governing equations are solved numerically using the interface-correction level set/ghost fluid method (ICLS/GFM) \citep{ICLS}. The equations are discretized in space by the finite volume method, and integrated in time using the second-order Adam-Bashforth scheme. Combining several computational techniques, including the fast pressure correction method \citep{Dodd_JCP_2014}, the GFM \citep{Fedkiw_JCP_1999}, and the fast Fourier transform (FFT) \citep{Schumann_JCP_1988}, discontinuous quantities are treated sharply at high efficiency. For the detailed algorithm and validations, we refer to  \cite{ICLS}. The source code is also publicly available on GitHub \citep{icls_code}.

@@

The simplest case of droplet self-assembly is identical drops approaching in quiescent flows, i.e.\ no external flow motions. In such a case, the remaining fluid parameters can be grouped into two non-dimensional numbers, i) the Laplace number $La = \tilde{\sigma} \tilde{\rho}_1 (2\tilde{R})/\tilde{\mu}_1^2$, where $\tilde{R}$ is the droplet radius; and ii) the osmotic-to-Laplace pressure ratio $\Pi=p_{os}/p$. \footnote[3]{The density and viscosity ratios between the drop and the carrier fluids are assumed to be unity. See Appendix \ref{app:numerics} for the detailed numerical setup.} $La$ relates surface tension to the viscous stress, which can also be expressed as $La = Re/Ca$. $\Pi$ indicates the magnitude of the uniform osmotic pressure $p_{os}$ due to depletion of the surfactant micelles, scaled by a reference Laplace pressure $p$ due to surface tension (cf.\ Eq.\ \eqref{Pi}). In the following, we assume $La = 2000$ and $\Pi = 1/8$ or $1/2$, corresponding to the limit $Ca \ll Re \ll 1 $ and conditions above the critical micelle concentrations (CMC, see dimensional analysis below) as in the experiments \citep{Shen_2016AS}. 
%This is reasonable since Re $\sim 10^{-3}$ and Ca $\sim 10^{-6}$ (see \citep{shen_thesis}, pp 63-64), and the osmotic pressure at surfactant concentrations above CMC is of similar order-of-magnitude to the Laplace pressure (see dimensional values below). The same setup is used throughout the section, except that the domain size is adjusted to accommodate more droplets when necessary.

@@

We note that $T_\pi$ is not a physical time scale (it is dimensionless). One possible definition for the depletion time scale is $\tilde{\tau}_\pi = \tilde{r}_s \tilde{\mu}_1 /(\tilde{R} \tilde{p}_{os})$, which can be rewritten as $T_\pi Ca \tilde{\tau}$, with a convection time scale $\tilde{\tau} =2\tilde{R}/\tilde{U}$ in the range of $0.1$ to $1$ s typically. By substitution of usual values of colloidal systems, e.g. $\tilde{r}_s = 1$ nm, $\tilde{\mu}_1 = 10^{-3}$ kg/m-s, $\tilde{R} = 10$ $\mu$m, and $\tilde{p}_{os} = 100$ Pa (corresponding to surfactant micelles concentration of 5 CMC, see  \cite{shen_thesis}, p112), the estimated time scale is 1 ns. Although $\tilde{\tau}_\pi$ can be amplified by increasing the viscosity of the suspending fluid or reducing the micelles concentration, its magnitude is so small that the approaching can be considered instantaneous. Therefore, in practical microfluidic devices such as those in   \cite{Shen_2016AS}, one cannot expect to detect the dynamical approaching process due to depletion forces. The droplets will appear either bound or separated, depending on the surfactant concentration and flow conditions.

@@

To further demonstrate the effect of the depletion force, we ``virtually'' assemble four to ten droplets under various initial configurations to form stable clusters as illustrated in Fig.\ \ref{fig: sph pack}. These clusters can be either $2D$ or $3D$, exhibiting different levels/kinds of symmetry. In our simulations, the shape of the cluster is solely determined by the initial droplet arrangement, in the absence of any disturbance or other driving forces. Permitting disturbances, such as vibrations or thermal noises, would eventually lead to the formation of ``rigid clusters", i.e.\ clusters that cannot be reshaped by a small amount of inter-droplet displacement. Analytically, the number of possible rigid clusters grows rapidly with the number of droplets ($N$). For example, there is only one possible rigid cluster for $N = 4$, while there are 259 possibilities for $N = 10$, for packing of $3N-6$ contacts  \citep{Holmes-Cerfon}. For brevity, we only illustrate two examples for $N = 4$ (including one planar cluster) and one example for $N = 5$ to 10 in Fig. \ref{fig: sph pack}.

We remind that the specific coordinates of the sphere packings bear no more significance than other possibilities in our simulations. They are arbitrarily chosen to illustrate the self-assembly due to the near field attraction. This is not the case for trapped equilibrium clusters, where less symmetric geometries are found to be favored by the entropic depletion force \citep{Meng2010,Klein2018}. \xx{Nor is it similar to colloidal particles interacting via short-range attracitve, long-range repulsive potentials, where complex phase transitions emerge depending on the competition of the interactions \citep{salr14,salr18}.} Here, the self-assembly is microfluidic-based, driven by the hydrodynamics rather than the minimization of free energy over long periods. We examine the effect of the flow next.

@@

When the droplets are carried by an external flow, their interactions are undoubtedly affected by the flow conditions, droplet-flow interactions and flow-induced droplet-droplet interactions \citep{Fouxon_2017}. To study these additional effects, we consider an elementary flow field, the wall-bounded simple shear flow, defined as $(u,v,w)=(0, \dot{\gamma}z,0)$ for $z \in [-L_z/2,L_z/2]$. Here, $v$ is the only non-zero velocity component, its magnitude varies linearly with the $z$ coordinate, and $\dot{\gamma}$ is the shear rate (see Fig.\  \ref{fig: shell}a). The presence of droplets will locally modify this flow field, which we sustain by enforcing opposite motions of two moving plates at $z= \pm L_z/2$.

Dating back to  \cite{Taylor501}, the deformation and motion of single or multiple droplet(s)/particle(s) have been studied extensively in simple shear flows \citep{lin_lee_sather_1970,batchelor_green_1972,batchelor_green_1972b,Zinchenko1983,Zinchenko1984,zurita-gotor_2007}. In the case of spherical particles/droplets, previously identified interaction modes include closed trajectories (particles rotate around each other in the vorticity plane), open-and-symmetric trajectories (particles return to their original $z$ positions after passing each other), and swapping trajectories (particles exchange the $z$ position after a binary encounter). The first two modes are generic features of a dilute suspension of particles or non-deforming/non-coalescing droplets, while the last mode arises when the particles are under relatively large geometric confinement. Conceivably, adding a near-field depletion force shall not alter these three modes as it is only activated at nearly touching, when the particles are already in a bound pair. What is yet to be explored, however, is when the gap between the two confining plates is smaller than the sum of the particle diameters (i.e.\ $L_z/D<2$), thus disabling the occurrence of the Batchelor-Green type of closed orbit.

@@

Fig.\ \ref{fig: shell}(a) illustrates various initial conditions of two touching droplets, corresponding to the nine cases listed in Table \ref{tbl: cases}, at $L_z/D=1.5$. Specifically, the initial positions of the second droplet is given in the spherical coordinates centered at the first one, where all cases are located in one quadrant-sphere as we do not distinguish between the two droplets (the rest are equivalent due to symmetry). Contrary to the less confined conditions where a bound pair would rotate indefinitely relative to each other under shear \citep{batchelor_green_1972,Zinchenko1984}, Fig.\ \ref{fig: shell} shows that the droplets tend to reside in the vorticity plane (i.e.\ the $yz$ plane) with a stable polar angle $\theta_\infty \approx 79$ deg. This is true even if the two droplets are not initially in the same vorticity plane (case 2,3,5,6,9), or if the second droplet is in the lower hemisphere relative to the first one (case 7,8,9). Particularly, in cases 8 and 9, the droplet pair first rotates clockwise, then slides along the wall, before finally reaching the stable orientation (see ESI Video 1). The anomalous trajectory is a clear evidence of the influence of the walls, which, together with the attractive depletion force, break the symmetry of the droplet binary interactions.

@@

\xx{In the following, clusters of three or four droplets are initialized to be in contact ($N-1$ contacts for $N$ droplets) and are released into different regions of a microfluidic channel to study their transport behavior. The production of these droplets is omitted, as the step-emulsifier is typically much smaller than the size of the channel, allowing for separation of the two processes \citep{shen_thesis,step-emuls}. We note that, although the droplets are already in a cluster initially, their relative rearrangement is still important as it determines the cluster morphology in the final state. The latter results primarily from the droplet-flow interaction and has direct consequence on the photonic properties of the droplet lattice, as we will discuss in details below.}

@@

As a control case, we simulate three droplets initially located near the bottom of the channel without any depletion forces. This is shown in the first column of Fig.\ \ref{fig: poi results}a, where the top and bottom panels illustrate the top and side views of the channel (see Fig.\ \ref{fig: chip}). The two snapshots are separated by $13.5$ convection time units (i.e.\ $\tilde{\tau}$). Clearly, the droplets quickly scatter as carried by the flow. In a previous work, we theoretically predicted the emergence of singlets and pairs of a dilute particle suspension due to weak particle-particle interactions \citep{Fouxon_2017}. This example illustrates the separation of a droplet cluster, enhanced by their initial proximity already at $N=3$, supporting our theoretical predictions.

@@

The above results confirm that the chain-like structure is indeed the predominant configuration of droplets bound by short-range depletion forces in the Poiseuille flow. Experimentally, this corresponds to strong diluting flows at the channel inlet, where long droplet strings are also observed further downstream (see  \cite{shen_thesis}, p137). More importantly, our simulations suggest that aligning of the droplets is a $3D$ shear-induced effect mediated by the confinement. The cross-stream migration of the droplets happens within a much shorter time span than any tangential rearrangement due to the dipolar interactions (cf.\  \cite{Diamant} and  \cite{Fouxon_2017}). This is one key difference between our $3D$ microfluidic channel and other $q2D$ devices.
